The Lipid Interactome:

### 2025-11-28 BioinformaticsApplicationNote/GG_2025_applicationNote.tex
Pushing latest formatting edit of manuscript

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The Lipid Interactome},
  pdfkeywords={Lipid-protein
interactions, Proteomics, Interactomics, Database, Multifunctionalized
lipids},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{The Lipid Interactome}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{An interactive and open access platform for exploring cellular
lipid-protein interactomes}
\author{Gaelen Guzman \and André Nadler \and Frank Stein \and Jeremy M.
Baskin \and Carsten Schultz \and Fikadu Tafesse}
\date{2025-02-04}

@@

Gaelen Guzman\textsuperscript{1,}*, André Nadler\textsuperscript{2},
Frank Stein\textsuperscript{3}, Jeremy M. Baskin\textsuperscript{4},
Carsten Schulz\textsuperscript{1}, Fikadu G. Tafesse\textsuperscript{1}

\textsuperscript{1} = Departments of Molecular Immunology and Immunology
and Chemical Physiology and Biochemistry, Oregon Health \& Science
University, Portland, OR, USA; \textsuperscript{2} = Max Plank Institute
of Molecular Cell Biology and Genetics, Dresden Germany;
\textsuperscript{3} = Proteomics Core Facility, European Molecular
Biology Laboratory, Heidelberg, Germany; \textsuperscript{4} =
Department of Chemistry and Chemical Biology and Weill Institute for
Cell and Molecular Biology, Cornell University, Ithaca, NY, USA; * =
Corresponding author:
\href{mailto:gaelen.guzman@lipidinteractome.org}{\nolinkurl{gaelen.guzman@lipidinteractome.org}}

@@

Lipid--protein interactions play essential roles in cellular signaling
and membrane dynamics, yet their systematic characterization has long
been hindered by the inherent biochemical properties of lipids. Recent
advances in functionalized lipid probes -- equipped with
photoactivatable crosslinkers, affinity handles, and photocleavable
protecting groups -- have enabled proteomics-based identification of
lipid interacting proteins with unprecedented specificity and
resolution. Despite the growing number of published lipid interactomes,
there remains no centralized effort to harmonize, compare, or integrate
these datasets.

@@

Historically, lipids have been considered merely the passive structural
components of cellular membranes. However, recent decades have revealed
that this diverse category of biomolecule plays essential and active
roles in the cell, ranging from modulating membrane dynamics,
determining organelle identity, serving as catabolites for energy
storage and metabolism, and bioactive signaling mediators for organellar
and intracellular signaling\textsuperscript{1--6}. Interactions with
proteins are essential to many of these crucial lipid functions -- and
yet, lipid--protein interactions are extremely challenging to study: The
small size and biophysical properties of lipids make conventional
proteomics approaches unsuitable for mapping lipid
interactors\textsuperscript{7}. Unlike proteins, which can be readily
tagged or immobilized for affinity-based interaction studies, lipids
require specialized chemical modifications to enable selective and
controlled interaction profiling\textsuperscript{8}.

Recent advances in synthetic, multifunctionalized lipid analogs have
revolutionized the field, enabling the systematic interrogation of lipid
interactomes\textsuperscript{8}. These probes incorporate strategically
designed functional groups that enable covalent crosslinking, affinity
enrichment, and controlled activation. Most commonly, these include: (1)
a diazirine group for photoactivatable crosslinking upon exposure to 355
nm ultraviolet (UV) light, stabilizing non-covalent and often
modest-affinity lipid--protein interactions; (2) a terminal alkyne,
permitting click chemistry-based enrichment; and, in some cases, (3) a
coumarin photocage, which shields the lipid probe from premature
enzymatic modification until it is selectively cleaved by 405 nm light
(a wavelength that does not induce diazirine
photolysis)\textsuperscript{7--10}. Together, these modifications enable
direct proteomic identification of lipid interactors, distinguishing
true binding partners from transient or non-specific
associations\textsuperscript{7,10,11}.

Numerous studies have already employed these lipid probes to elucidate
the interactomes of key bioactive lipids, including diacylglycerol,
sphingosine, sphinganine, phosphatidic acid, and
phosphatidylethanolamine, N-acylphosphatidylethanolamine, and
phosphatidyl alcohol\textsuperscript{9,11--16}. These initial studies
have followed a comparative proteomics workflow, wherein crosslinked
(``+UV'') and non-crosslinked (``-UV'') samples are analyzed side by
side to minimize artifacts and confirm
specificity\textsuperscript{7,10}. However, despite the growing number
of lipid interactome datasets, no centralized resource has existed to
compile, compare, and standardize these findings.

To address this gap, we present the
\href{https://www.lipidinteractome.org}{Lipid Interactome} -- a curated
repository designed to facilitate the comparative analysis of
lipid--protein interactions from multiple studies. This resource
consolidates proteomics data from multifunctional lipid probes, enabling
researchers to efficiently explore and contextualize lipid-binding
proteins across different lipid species and cellular models. Currently,
the repository hosts interactomes derived from trifunctionalized
phosphatidic acid, phosphatidylethanolamine, phosphatidylinositol
\emph{tris}phosphate, N-acylphosphatidylethanolamine, phosphatidyl
alcohol, sphingosine, and sphinganine in HeLa and Huh7
cells\textsuperscript{11,13--16}. Additionally, it includes control
datasets from two fatty acid probes to distinguish specific from
non-specific lipid-binding proteins\textsuperscript{12}. We additionally
intend to expand this resource to include more complex experimental
designs as they are published (e.g.~Disease versus Healthy Control or
Infection versus Mock).

@@

Each publication contributing to the repository is provided with a
dedicated results page, allowing users to examine study findings in
detail. These pages include interactive~Volcano,~Ranked-Order, and~MA
plots, which enable visualization of differential protein enrichment
across experimental conditions. The plots are rendered using
the~Plotly~and~htmlwidgetsR packages, ensuring a responsive, client-side
experience that supports zooming, hovering, image export, and data
filtering\textsuperscript{17,18}. For transparency and reproducibility,
all datasets associated with a given study are available for direct
download.

@@

Where study design permits, data has been integrated into an
interactive~Shiny~application, allowing users to perform direct (though
qualitative)~probe-to-probe
comparisons~(\href{https://www.lipidinteractome.org/lipidprobe/enrichedhitscomparison}{Probe~vs~Probe~Comparisons}).
This tool enables researchers to explore the similarity of lipid
interactomes across experimental conditions via~linear regression
analysis, highlighting proteins that are significantly enriched in one
dataset, both datasets, or neither. By providing an intuitive framework
for interactome comparison, this feature aids in the identification of
promising targets for further investigation.

@@

While there are numerous ways to quantify the mass spectrometry (MS)
data, the majority of studies included in our repository have utilized
the same (or highly similar) quantitative MS pipeline using
data-dependent acquisition techniques, such as Tandem Mass
Tagging\textsuperscript{19}. These datasets include those reported by
Farley et al.~(2024), Farley et al.~(2024), and Thomas et
al.~(2025)\textsuperscript{11--13}. Some studies have also featured
quantitative MS analysis using stable isotope labeling of amino acids in
cell culture (SILAC), which also enables extraction of key data
described below\textsuperscript{14,15}. Correspondingly, the ideal data
structure for incorporation into our repository will include the
following:

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1486}}
  >{\centering\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1081}}
  >{\centering\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.3514}}
  >{\centering\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1216}}
  >{\centering\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1622}}
  >{\centering\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1081}}@{}}
\caption{\textbf{Example data table for smooth incorporation into Lipid
Interactome.} \textbf{\emph{Column descriptions:}} \emph{Gene name}: The
identifier for genes identified in the dataset (or Accession Number).
\emph{Uniprot accession}: Stable identifiers of UniProtKB entries;
integrates species and isoform data on the identified protein.
\emph{Enrichment value}: Fold change of Experimental versus Control
fold-change, typically log2-transformed -- ideally the product of TMT or
other quantitative mass spectrometry methods. \emph{p value}: the
results of an ANOVA or LIMMA analysis, corrected for multiple hypothesis
testing. \emph{AveExpr}: An averaged value of the m/z ion intensity of
the Experimental condition and Control condition used for the enrichment
value (used to prepare MA plots for quality control). \emph{FDR}:
False-discovery rate of protein
identification.}\label{tbl-1}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Gene name
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\href{https://www.uniprot.org/help/accession_numbers}{Uniprot accession}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
logFC
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
AveExpr
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDR
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Gene name
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\href{https://www.uniprot.org/help/accession_numbers}{Uniprot accession}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
logFC
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
AveExpr
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDR
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
GeneA & P80723 & 1.2 & 0.001 & 1.0x10\textsuperscript{18} & 0.0001 \\
GeneB & P32004 & 0.1 & 0.6 & 1.0x10\textsuperscript{5} & 0.005 \\
\ldots{} & \ldots{} & \ldots{} & \ldots{} & \ldots{} & \ldots{} \\
\end{longtable}

\textbf{We wish to emphasize that we will collaborate with data
providers to prepare visualizations which best suit their datasets --
the example dataset above merely depicts an ideal case which allows for
easy incorporation and cross-study comparison.}

Additionally, we are working on providing visualization functionality
for studies which utilized Peptide Spectral Matching (PSM) to report
enrichment toward lipid probes. Similarly, data-independent acquisitio
(DIA) is a quickly emerging alternative to data-dependent acquisition
(DDA; e.g.~TMT)\textsuperscript{20,21}; as with PSM datasets, we are
working to incorporate DIA data formats.

@@

Designed as a continuously expanding repository, the Lipid Interactome
will evolve alongside advances in lipid probe technologies and
interactomics. Future iterations will incorporate validation studies for
each lipid probe, offering researchers a dedicated space to compare
findings across experimental contexts. This functionality will not only
reduce redundancy in experimental design but also strengthen the
reproducibility and biological interpretation of lipid interactome
research. To support this vision, we invite researchers to contribute
their datasets via our
\href{https://lipidinteractome.org/contactus/datasubmission}{Data
Submission page}, fostering a collaborative and transparent approach to
lipid biology.

@@

\phantomsection\label{refs}
\begin{CSLReferences}{0}{0}
\bibitem[\citeproctext]{ref-barischMembraneDamageRepair2023}
\CSLLeftMargin{1. }%
\CSLRightInline{Barisch, C., Holthuis, J. C. M. \& Cosentino, K.
\href{https://doi.org/10.1515/hsz-2022-0321}{Membrane damage and repair:
A thin line between life and death}. \emph{Biological Chemistry}
\textbf{404}, 467--490 (2023).}

\bibitem[\citeproctext]{ref-vanmeerMembraneLipidsWhere2008a}
\CSLLeftMargin{2. }%
\CSLRightInline{van Meer, G., Voelker, D. R. \& Feigenson, G. W.
\href{https://doi.org/10.1038/nrm2330}{Membrane lipids: Where they are
and how they behave}. \emph{Nature Reviews Molecular Cell Biology}
\textbf{9}, 112--124 (2008).}

\bibitem[\citeproctext]{ref-holthuisLipidLandscapesPipelines2014a}
\CSLLeftMargin{3. }%
\CSLRightInline{Holthuis, J. C. M. \& Menon, A. K.
\href{https://doi.org/10.1038/nature13474}{Lipid landscapes and
pipelines in membrane homeostasis}. \emph{Nature} \textbf{510}, 48--57
(2014).}

\bibitem[\citeproctext]{ref-niphakisGlobalMapLipidBinding2015a}
\CSLLeftMargin{4. }%
\CSLRightInline{Niphakis, M. J. \emph{et al.}
\href{https://doi.org/10.1016/j.cell.2015.05.045}{A {Global Map} of
{Lipid-Binding Proteins} and {Their Ligandability} in {Cells}}.
\emph{Cell} \textbf{161}, 1668--1680 (2015).}

\bibitem[\citeproctext]{ref-chernomordikProteinLipidInterplayFusion2003}
\CSLLeftMargin{5. }%
\CSLRightInline{Chernomordik, L. V. \& Kozlov, M. M.
\href{https://doi.org/10.1146/annurev.biochem.72.121801.161504}{Protein-{Lipid
Interplay} in {Fusion} and {Fission} of {Biological Membranes}}.
\emph{Annual Review of Biochemistry} \textbf{72}, 175--207 (2003).}

\bibitem[\citeproctext]{ref-simonsLipidRaftsSignal2000a}
\CSLLeftMargin{6. }%
\CSLRightInline{Simons, K. \& Toomre, D.
\href{https://doi.org/10.1038/35036052}{Lipid rafts and signal
transduction}. \emph{Nature Reviews Molecular Cell Biology} \textbf{1},
31--39 (2000).}

\bibitem[\citeproctext]{ref-schultzChemicalToolsLipid2023}
\CSLLeftMargin{7. }%
\CSLRightInline{Schultz, C.
\href{https://doi.org/10.1021/acs.accounts.2c00851}{Chemical {Tools} for
{Lipid Cell Biology}}. \emph{Accounts of Chemical Research} \textbf{56},
1168--1177 (2023).}

\bibitem[\citeproctext]{ref-schultzFlashClickMultifunctionalized2022a}
\CSLLeftMargin{8. }%
\CSLRightInline{Schultz, C., Farley, S. E. \& Tafesse, F. G.
\href{https://doi.org/10.1021/jacs.2c02705}{{`{Flash} \& {Click}'}:
{Multifunctionalized Lipid Derivatives} as {Tools To Study Viral
Infections}}. \emph{Journal of the American Chemical Society}
\textbf{144}, 13987--13995 (2022).}

\bibitem[\citeproctext]{ref-hoglingerTrifunctionalLipidProbes2017}
\CSLLeftMargin{9. }%
\CSLRightInline{Höglinger, D. \emph{et al.}
\href{https://doi.org/10.1073/pnas.1611096114}{Trifunctional lipid
probes for comprehensive studies of single lipid species in living
cells}. \emph{Proceedings of the National Academy of Sciences}
\textbf{114}, 1566--1571 (2017).}

\bibitem[\citeproctext]{ref-haberkantFatFabulousBifunctional2014}
\CSLLeftMargin{10. }%
\CSLRightInline{Haberkant, P. \& Holthuis, J. C. M.
\href{https://doi.org/10.1016/j.bbalip.2014.01.003}{Fat \& fabulous:
{Bifunctional} lipids in the spotlight}. \emph{Biochimica et Biophysica
Acta - Molecular and Cell Biology of Lipids} \textbf{1841}, 1022--1030
(2014).}

\bibitem[\citeproctext]{ref-farleyTrifunctionalSphinganineNew2024a}
\CSLLeftMargin{11. }%
\CSLRightInline{Farley, S., Stein, F., Haberkant, P., Tafesse, F. G. \&
Schultz, C.
\href{https://doi.org/10.1021/acschembio.3c00554}{Trifunctional
{Sphinganine}: {A New Tool} to {Dissect Sphingolipid Function}}.
\emph{ACS Chemical Biology} \textbf{19}, 336--347 (2024).}

\bibitem[\citeproctext]{ref-farleyTrifunctionalFattyAcid2024}
\CSLLeftMargin{12. }%
\CSLRightInline{Farley, S. E. \emph{et al.}
\href{https://doi.org/10.1039/D4CC00974F}{Trifunctional fatty acid
derivatives: The impact of diazirine placement}. \emph{Chemical
Communications} \textbf{60}, 6651--6654 (2024).}

\bibitem[\citeproctext]{ref-thomasTrifunctionalLipidDerivatives2025}
\CSLLeftMargin{13. }%
\CSLRightInline{Thomas, A. \emph{et al.} Trifunctional lipid
derivatives: {PE}'s mitochondrial interactome. \emph{Chemical
Communications} 10.1039.D4CC03599B (2025)
doi:\href{https://doi.org/10.1039/D4CC03599B}{10.1039/D4CC03599B}.}

\bibitem[\citeproctext]{ref-chiuPhotoaffinityLabelingReveals2025}
\CSLLeftMargin{14. }%
\CSLRightInline{Chiu, D.-C., Lin, H. \& Baskin, J. M. Photoaffinity
{Labeling Reveals} a {Role} for the {Unusual Triply Acylated
Phospholipid N-Acylphosphatidylethanolamine} in {Lactate Homeostasis}.
(2025)
doi:\href{https://doi.org/10.26434/chemrxiv-2025-j8nqz}{10.26434/chemrxiv-2025-j8nqz}.}

\bibitem[\citeproctext]{ref-yuChemoproteomicsApproachProfile2022}
\CSLLeftMargin{15. }%
\CSLRightInline{Yu, W., Lin, Z., Woo, C. M. \& Baskin, J. M.
\href{https://doi.org/10.1021/acschembio.1c00584}{A {Chemoproteomics
Approach} to {Profile Phospholipase D-Derived Phosphatidyl Alcohol
Interactions}}. \emph{ACS Chemical Biology} \textbf{17}, 3276--3283
(2022).}

\bibitem[\citeproctext]{ref-mullerSynthesisCellularLabeling2021}
\CSLLeftMargin{16. }%
\CSLRightInline{Muller, R., Kojic, A., Citir, M. \& Schultz, C.
\href{https://doi.org/10.1002/anie.202103599}{Synthesis and {Cellular
Labeling} of {Multifunctional Phosphatidylinositol Bis}- and
{Trisphosphate Derivatives}}. \emph{Angewandte Chemie International
Edition} \textbf{60}, 19759--19765 (2021).}

\bibitem[\citeproctext]{ref-plotly}
\CSLLeftMargin{17. }%
\CSLRightInline{Inc., P. T. Collaborative data science. (2015).}

\bibitem[\citeproctext]{ref-ramnathvaidyanathanHtmlwidgetsHTMLWidgets2015}
\CSLLeftMargin{18. }%
\CSLRightInline{Ramnath Vaidyanathan \emph{et al.} Htmlwidgets: {HTML
Widgets} for {R}. (2015).}

\bibitem[\citeproctext]{ref-thompsonTandemMassTags2003}
\CSLLeftMargin{19. }%
\CSLRightInline{Thompson, A. \emph{et al.}
\href{https://doi.org/10.1021/ac0262560}{Tandem {Mass Tags}: {A Novel
Quantification Strategy} for {Comparative Analysis} of {Complex Protein
Mixtures} by {MS}/{MS}}. \emph{Analytical Chemistry} \textbf{75},
1895--1904 (2003).}

\bibitem[\citeproctext]{ref-birhanuMassSpectrometrybasedProteomics2023}
\CSLLeftMargin{20. }%
\CSLRightInline{Birhanu, A. G.
\href{https://doi.org/10.1186/s12014-023-09424-x}{Mass
spectrometry-based proteomics as an emerging tool in clinical
laboratories}. \emph{Clinical Proteomics} \textbf{20}, (2023).}

\bibitem[\citeproctext]{ref-boysClinicalApplicationsMass2023}
\CSLLeftMargin{21. }%
\CSLRightInline{Boys, E. L., Liu, J., Robinson, P. J. \& Reddel, R. R.
\href{https://doi.org/10.1002/pmic.202200238}{Clinical applications of
mass spectrometry-based proteomics in cancer: {Where} are we?}
\emph{PROTEOMICS} \textbf{23}, 2200238 (2023).}

\end{CSLReferences}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The Lipid Interactome:},
  pdfkeywords={Lipid-protein
interactions, Proteomics, Interactomics, Database, Multifunctionalized
lipids},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{The Lipid Interactome:}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{An interactive and open access platform for exploring cellular
lipid-protein interactions}
\author{Gaelen Guzman \and André Nadler \and Frank Stein \and Jeremy M.
Baskin \and Carsten Schultz \and Fikadu Tafesse}
\date{2025-02-04}

@@

Gaelen Guzman\textsuperscript{1,}*, André Nadler\textsuperscript{3},
Frank Stein\textsuperscript{4}, Jeremy M. Baskin\textsuperscript{5},
Carsten Schulz\textsuperscript{2}, Fikadu G.
Tafesse\textsuperscript{1,2}

\textsuperscript{1}Department of Molecular Immunology \& Immunology,
Oregon Health \& Science University, Portland, OR, USA;
\textsuperscript{2}Chemical Physiology and Biochemistry, Oregon Health
\& Science University, Portland, OR, USA; \textsuperscript{3}Max Plank
Institute of Molecular Cell Biology and Genetics, Dresden Germany;
\textsuperscript{4}Proteomics Core Facility, European Molecular Biology
Laboratory, Heidelberg, Germany; \textsuperscript{5}Department of
Chemistry and Chemical Biology and Weill Institute for Cell and
Molecular Biology, Cornell University, Ithaca, NY, USA; *Corresponding
author:
\href{mailto:gaelen.guzman@lipidinteractome.org}{\nolinkurl{gaelen.guzman@lipidinteractome.org}}

@@

Lipid-protein interactions play essential roles in cellular signaling
and membrane dynamics, yet their systematic characterization has long
been hindered by the inherent biochemical properties of lipids. Recent
advances in functionalized lipid probes -- equipped with
photoactivatable crosslinkers, affinity handles, and photocleavable
protecting groups -- have enabled proteomics-based identification of
lipid interacting proteins with unprecedented specificity and
resolution. Despite the growing number of published lipid interactomes,
there remains no centralized effort to harmonize, compare, or integrate
these datasets.

@@

Historically, lipids have been considered merely the passive structural
components of cellular membranes. However, recent decades have revealed
that this diverse category of biomolecule plays essential and active
roles in the cell, ranging from modulating membrane dynamics,
determining organelle identity, serving as catabolites for energy
storage and metabolism, and bioactive signaling mediators for organellar
and intracellular signaling{[}@barischMembraneDamageRepair2023;
@vanmeerMembraneLipidsWhere2008a;
@holthuisLipidLandscapesPipelines2014a;
@niphakisGlobalMapLipidBinding2015a;
@chernomordikProteinLipidInterplayFusion2003;
@simonsLipidRaftsSignal2000a{]}. Interactions with proteins are
essential to many of these crucial lipid functions -- and yet,
lipid--protein interactions are extremely challenging to study: The
small size and biophysical properties of lipids make conventional
proteomics approaches unsuitable for mapping lipid
interactors{[}@schultzChemicalToolsLipid2023{]}. Unlike proteins, which
can be readily tagged or immobilized for affinity-based interaction
studies, lipids require specialized chemical modifications to enable
selective and controlled interaction
profiling{[}@schultzFlashClickMultifunctionalized2022a{]}.

Recent advances in synthetic, multifunctionalized lipid analogs have
revolutionized the field, enabling the systematic interrogation of lipid
interactomes{[}@schultzFlashClickMultifunctionalized2022a{]}. These
probes incorporate strategically designed functional groups that enable
covalent crosslinking, affinity enrichment, and controlled activation.
Most commonly, these include: (1) a diazirine group for photoactivatable
crosslinking upon exposure to 355 nm ultraviolet (UV) light, stabilizing
non-covalent and often modest-affinity lipid--protein interactions; (2)
a terminal alkyne, permitting click chemistry-based enrichment; and, in
some cases, (3) a coumarin photocage, which shields the lipid probe from
premature enzymatic modification until it is selectively cleaved by 405
nm light (a wavelength that does not induce diazirine
photolysis){[}@schultzFlashClickMultifunctionalized2022a;
@schultzChemicalToolsLipid2023; @hoglingerTrifunctionalLipidProbes2017;
@haberkantFatFabulousBifunctional2014{]}. Together, these modifications
enable direct proteomic identification of lipid interactors,
distinguishing true binding partners from transient or non-specific
associations{[}@haberkantFatFabulousBifunctional2014;
@schultzChemicalToolsLipid2023;
@farleyTrifunctionalSphinganineNew2024a{]}.

Numerous studies have already employed these lipid probes to elucidate
the interactomes of key bioactive lipids, including diacylglycerol,
sphingosine, sphinganine, phosphatidic acid, and
phosphatidylethanolamine, N-acylphosphatidylethanolamine, and
phosphatidyl alcohol{[}@hoglingerTrifunctionalLipidProbes2017;
@farleyTrifunctionalSphinganineNew2024a;
@farleyTrifunctionalFattyAcid2024;
@thomasTrifunctionalLipidDerivatives2025;
@chiuPhotoaffinityLabelingReveals2025;
@yuChemoproteomicsApproachProfile2022;
@mullerSynthesisCellularLabeling2021{]}. These initial studies have
followed a comparative proteomics workflow, wherein crosslinked
(``+UV'') and non-crosslinked (``-UV'') samples are analyzed side by
side to minimize artifacts and confirm
specificity{[}@haberkantFatFabulousBifunctional2014;
@schultzChemicalToolsLipid2023{]}. However, despite the growing number
of lipid interactome datasets, no centralized resource has existed to
compile, compare, and standardize these findings.

To address this gap, we present the
\href{https://www.lipidinteractome.org}{Lipid Interactome} -- a curated
repository designed to facilitate the comparative analysis of
lipid--protein interactions from multiple studies. This resource
consolidates proteomics data from multifunctional lipid probes, enabling
researchers to efficiently explore and contextualize lipid-binding
proteins across different lipid species and cellular models. Currently,
the repository hosts interactomes derived from trifunctionalized
phosphatidic acid, phosphatidylethanolamine, phosphatidylinositol
\emph{tris}phosphate, N-acylphosphatidylethanolamine, phosphatidyl
alcohol, sphingosine, and sphinganine in HeLa and Huh7
cells{[}@thomasTrifunctionalLipidDerivatives2025;
@mullerSynthesisCellularLabeling2021;
@chiuPhotoaffinityLabelingReveals2025;
@yuChemoproteomicsApproachProfile2022;
@farleyTrifunctionalSphinganineNew2024a{]}. Additionally, it includes
control datasets from two fatty acid probes to distinguish specific from
non-specific lipid-binding
proteins{[}@farleyTrifunctionalFattyAcid2024{]}. We additionally intend
to expand this resource to include more complex experimental designs as
they are published (e.g.~Disease versus Healthy Control or Infection
versus Mock).

@@

Each publication contributing to the repository is provided with a
dedicated results page, allowing users to examine study findings in
detail. These pages include interactive~Volcano,~Ranked-Order, and~MA
plots, which enable visualization of differential protein enrichment
across experimental conditions. The plots are rendered using
the~Plotly~and~htmlwidgetsR packages, ensuring a responsive, client-side
experience that supports zooming, hovering, image export, and data
filtering{[}@plotly; @ramnathvaidyanathanHtmlwidgetsHTMLWidgets2015{]}.
For transparency and reproducibility, all datasets associated with a
given study are available for direct download.

@@

Where study design permits, data has been integrated into an
interactive~Shiny~application, allowing users to perform direct (though
qualitative)~probe-to-probe
comparisons~(\href{https://www.lipidinteractome.org/probevsprobecomparison/enrichedhitscomparison}{Probe~vs~Probe~Comparisons}).
This tool enables researchers to explore the similarity of lipid
interactomes across experimental conditions via~linear regression
analysis, highlighting proteins that are significantly enriched in one
dataset, both datasets, or neither. By providing an intuitive framework
for interactome comparison, this feature aids in the identification of
promising targets for further investigation.

@@

While there are numerous ways to quantify the mass spectrometry (MS)
data, the majority of studies included in our repository have utilized
the same (or highly similar) quantitative MS pipeline using
data-dependent acquisition techniques, such as Tandem Mass
Tagging{[}@thompsonTandemMassTags2003{]}. These datasets include those
reported by Farley et al.~(2024), Farley et al.~(2024), and Thomas et
al.~(2025){[}@farleyTrifunctionalSphinganineNew2024a;
@farleyTrifunctionalFattyAcid2024;
@thomasTrifunctionalLipidDerivatives2025{]}. Some studies have also
featured quantitative MS analysis using stable isotope labeling of amino
acids in cell culture (SILAC), which also enables extraction of key data
described below{[}@chiuPhotoaffinityLabelingReveals2025;
@yuChemoproteomicsApproachProfile2022{]}. Correspondingly, the ideal
data structure for incorporation into our repository will contain the
information shown in Table~\ref{tbl-dataStructure}.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1486}}
  >{\centering\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1081}}
  >{\centering\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.3514}}
  >{\centering\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1216}}
  >{\centering\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1622}}
  >{\centering\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1081}}@{}}
\caption{\textbf{Example data table for smooth incorporation into Lipid
Interactome.} \textbf{\emph{Column descriptions:}} \emph{Gene name}: The
identifier for genes identified in the dataset (or Accession Number).
\emph{Uniprot accession}: Stable identifiers of UniProtKB entries;
integrates species and isoform data on the identified protein.
\emph{Enrichment value}: Fold change of Experimental versus Control
fold-change, typically log2-transformed -- ideally the product of TMT or
other quantitative mass spectrometry methods. \emph{p value}: the
results of an ANOVA or LIMMA analysis, corrected for multiple hypothesis
testing. \emph{AveExpr}: An averaged value of the m/z ion intensity of
the Experimental condition and Control condition used for the enrichment
value (used to prepare MA plots for quality control). \emph{FDR}:
False-discovery rate of protein
identification.}\label{tbl-dataStructure}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Gene name
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\href{https://www.uniprot.org/help/accession_numbers}{Uniprot accession}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
logFC
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
AveExpr
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDR
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Gene name
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\href{https://www.uniprot.org/help/accession_numbers}{Uniprot accession}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
logFC
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
p value
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
AveExpr
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDR
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
GeneA & P80723 & 1.2 & 0.001 & 1.0x10\textsuperscript{18} & 0.0001 \\
GeneB & P32004 & 0.1 & 0.6 & 1.0x10\textsuperscript{5} & 0.005 \\
\ldots{} & \ldots{} & \ldots{} & \ldots{} & \ldots{} & \ldots{} \\
\end{longtable}

\textbf{We wish to emphasize that we will collaborate with data
providers to prepare visualizations which best suit their datasets --
the example dataset above merely depicts an ideal case which allows for
easy incorporation and cross-study comparison. As proteomics techniques
have improved rapidly in recent years, we have done our best to
incorporate all data structures into this repository.}

Additionally, we are working on providing visualization functionality
for studies which utilized Peptide Spectral Matching (PSM) to report
enrichment toward lipid probes. Similarly, data-independent acquisitio
(DIA) is a quickly emerging alternative to data-dependent acquisition
(DDA; e.g.~TMT) {[}@birhanuMassSpectrometrybasedProteomics2023;
@boysClinicalApplicationsMass2023{]}; as with PSM datasets, we are
working to incorporate DIA data formats.

At present, we are accepting datasets as analyzed by the original
researchers, including protein identifications, enrichment
quantification, and the results of statistical testing. Summaries of
both source and data wrangling can be found in the \emph{Study
Information} dropdown section of each data page. Further, the raw data
(as provided directly by authors or pulled from publicly-available
datasts) and data wrangling scripts are available for download in the
same dropdown tabs.

@@

Designed as a continuously expanding repository, the Lipid Interactome
will evolve alongside advances in lipid probe technologies and
interactomics. Ongoing efforts seek to incorporate enhanced scalability
of this site, more robust data storage methods, and automated data
submission. These functionalities will improve the utility of this
repository, enabling applications such as machine learning for pattern
recognition as the site grows. Future iterations will incorporate
validation studies for each lipid probe, offering researchers a
dedicated space to compare findings across experimental contexts. This
functionality will not only reduce redundancy in experimental design but
also strengthen the reproducibility and biological interpretation of
lipid interactome research.

To support this vision, we invite researchers to contribute their
datasets via our
\href{https://lipidinteractome.org/contactus/datasubmission}{Data
Submission page}. Using the FAIR data principles, we hope to foster a
collaborative and transparent approach to lipid biology.

@@

\phantomsection\label{refs}
\begin{CSLReferences}{0}{1}
\end{CSLReferences}
